---
title: "Intestine promoter binding"
author: "David C. King and Ambika"
date: "11/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(GenomicRanges)
```

## Goal

This analysis is looking for TFs that may explain intestine expression in the absence of ELT-2.

## Promoters

Genes which fall under different classes of intestine enrichment in Williams et al. 2023
are assigned promoters gene 5' -1Kb/200bp. 

The dataset has each promoter in the genome annotated with intestine expression DESeq information, as well as ELT-2 binding status gotten from Kudron et al. 2018.

```{r promoters}
promoters_df = readRDS('DATA/promoter_data.rds')
promoters_df %>% head()

```

### Promoters by intestine enrichment status



```{r enrichment_depletion}
LE.enriched = promoters_df %>% filter(embryo_int_exp != "depleted")
L1.enriched = promoters_df %>% filter(L1_int_exp != "depleted")
L3.enriched = promoters_df %>% filter(L3_int_exp != "depleted")
promoters_gr = makeGRangesFromDataFrame(L1.enriched, keep.extra.columns = T, seqinfo=Seqinfo(genome="ce11"))

```

```{r other-TFs}
TF.files = list.files('summit/modENCODEtracks/data/orig/L1',
                      full.names =TRUE,
                      pattern='*.bed')
factornames = lapply(str_split(TF.files, "[/.]"), function(x) x[[6]]) %>% unlist()
names(TF.files) = factornames

all.grs= lapply(TF.files, 
  function(fname) {
    factorname = str_split(fname, "[/.]")[[1]][6]
    data=read.table(fname)
    gr = makeGRangesFromDataFrame(data, keep.extra.columns = T, 
                                  seqnames.field = "V1",
                                  start.field = "V2",
                                  end.field = "V3", seqinfo=Seqinfo(genome="ce11"))
    return(gr)
  })
```

```{r}


for(TFname in names(all.grs))
{
  
  olap=countOverlaps(promoters_gr,all.grs[[TFname]])
  
  df = mcols(promoters_gr)# pull dataframe from GRanges metadata
  df[[TFname]] = olap
  
   mcols(promoters_gr) <- df # reassign dataframe to GRanges
   
}

data = as.data.frame(mcols(promoters_gr))
data =data[,19:ncol(data)]
colSums(as.data.frame(data))
 
mtx_x <- matrix(unlist(data), ncol = 76, byrow = TRUE)
heatmap(mtx_x)
 
```
 